Wait-free Queues with Polygarithmic Step Complexity

### 2022-04-26 article/psuedocode.tex
apply meeting 4-5 feedbacks

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList} blocks}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores the index of the superblock of some block in \tt{blocks} whose \tt{group} field is \tt{i}.}
\end{itemize}

@@

\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}

@@

\Statex
\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \tt{<r\sub{deq}, b\sub{deq}>=} \tt{leaf.}\Call{IndexDeq}{leaf.size, 1}
\Statex \cmt{\tt{r} is the rank among all dequeues of the dequeue and \tt{b\sub{deq}} is the index of the block in the root containing the dequeue.}
\State \tt{<r\sub{enq}, b\sub{r\sub{enq}}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
\Statex \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue and \tt{b\sub{deq}} is the index of the block of it in the blocklist. If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
\If{\tt{r\sub{enq}==-1}}
\State \tt{output= null}
\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
%\cmt{all the operations in the block containing the dequeue are finished.}
\State \tt{leaf.last\sub{done}= b\sub{deq}}
\EndIf
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{r\sub{enq}, b\sub{r\sub{enq}}}} \cmt{getting the reponse's \tt{element}.}
\State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
\State \tt{root.blocks[b\sub{r\sub{enq}}].num\sub{finished}.inc()}
\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
\State \tt{leaf.last\sub{done}= b\sub{deq}}
\ElsIf{\tt{root.blocks[b\sub{r}].num\sub{finished}==root.blocks[b\sub{r}].num}}
\State \tt{leaf.last\sub{done}= b\sub{r\sub{enq}}}
\EndIf
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.}
\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
\State \Return \tt{<r\sub{enq}, root.blocks.get(enq, r\sub{enq}).order>}
\EndIf
\EndFunction{FindResponse}

@@

\Function{void}{Propagate()}{}
\If{\bf{not} \tt{this.}\Call{Refresh()}{}}
\State \tt{this.}\Call{Refresh()}{} \cmt{Lemma Double Refresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}} \cmt{To check a node is the root we can check its index is 0.} 
\State \tt{this.parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= size}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= this.\Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}}s.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{root.blocks.tryAppend(new)}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, h, h+1}}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. It might fell sleep before increasing.}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{feild}. Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
%\State \Return \tt{result block's index}
\EndFunction{BSearch}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into \tt{this.blocks[i]}. Returns the created block as well as values read from each childnum\sub{propagated} field. The values are used for incrementing children's num\sub{propagated}s if the block was appended to \tt{this.blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\State \tt{newBlock.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.size} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{n\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

@@

\State \Return \tt{b, np\sub{left}, np\sub{right}}
\EndFunction{CreateBlock}

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{n.blocks[b]} contains \tt{i}th enqueue in the node.}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
\If{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
\State \tt{subBlock= left\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}}
\State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}}
\State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\EndIf
\EndFunction{GetEnq}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{i}th dequeue in the node is in \tt{blocks[b]}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the order in the root of $i$th dequeue in the $b$th block of node n among dequeues.}
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
\State \tt{superBlock= \Call{BSearch}{parent, sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
\If{\tt{dir {\keywordfont is} left}}
\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
\Else \State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq} + blocks[superBlock].sum\sub{deq-left}} \cmt{consider dequeues from the right child}
\EndIf
\State \Return\Call{IndexDeq}{this.parent, superBlock, i}

\EndFunction{Index}

@@

\Function{void}{Append}{\sl{block} blk} 
\State \tt{size+=1} \label{appendEnd} \cmt{Lines \ref{appendStart} to \ref{appendEnd} are done by one process at time.}
\State \tt{blk.group= size} \label{appendStart} \cmt{Append is only called by the owener of the leaf.}
\State \tt{blocks[size]= blk} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

@@

\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
\State\Return \tt{blocks[b].element}
\EndFunction{GetEnq}

@@

\begin{algorithm}
\caption{Root}
\begin{algorithmic}[1]
\setcounter{ALG@line}{600}

\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} 
\State\Return \tt{<i, b>}
\EndFunction{IndexDeq}

\begin{algorithm}
\caption{BlockList}
\begin{algorithmic}[1]
\setcounter{ALG@line}{700}

\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}. Initially  empty, when \tt{blocks.tryAppend(b)} returns true \tt{b} is appended to the end of the list and \tt{blocks[i]} returns $i$th block in the \tt{blocks}. If some instance of \tt{blocks.tryAppend(b)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}

\Statex
\Statex $\diamondsuit$ \tt{\sl{PBRT implementation}}
\Function{boolean}{TryAppend}{\sl{block} blk} 
\State \tt{TODO}
\EndFunction{TryAppend}

\Statex
\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
\Function{boolean}{TryAppend}{\sl{block} blk} 
\State \tt{TODO}
\EndFunction{TryAppend}

@@

\Statex $\blacktriangleright$ \tt{\sl{PRBTree[rootBlock]}}
  \Statex \textsf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
  \tt{append(b, key)} returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.

\Function{void}{RBTAppend}{block b} \cmt{\textsf{adds block b to the \tt{root.blocks}}}
\State \tt{step= root.size}
\If{\tt{step\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root. Used in lemma's using the size of the PBRT.}
\State \tt{Help()}
\State \tt{CollectGarbage()}
\EndIf
\State \tt{b.num\sub{finished}= 0}
\State \Return \tt{root.blocks.append(b, b.order)}
\EndFunction{RBTAppend}
\Statex

\Function{void}{Help}{}\cmt{Helps pending operations}
\For{\tt{leaf l} \bf{in leaves}}\cmt{\it{if the tree is implemented with an array we can iterate over the second half of the array.}}
\State{\tt{last= l.size-1}}
\cmt{\tt{l.blocks[last]} can not be \tt{null} because of lines \ref{appendStart}-\ref{appendEnd}.}
\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
\State \tt{goto \ref{deqRest} with these values <>} \cmt{run \tt{Dequeue()}  for \tt{l.ops[last]} after Propagate(). \it{TODO}}
\State \tt{l.responses[last]= response}
\EndIf
\EndFor
\EndFunction{Help}
\Statex

@@

\Function{Block}{FindMostRecentDone}{b}
\For{\tt{leaf l} \bf{in leaves}}
\State\tt{max= Max(l.maxOld, max)}
\EndFor
\State\Return\tt{max} \cmt{This snapshot suffies.}
\EndFunction{findYoungestOld}

@@

\Function{response}{FallBack}{op i} \cmt{\it{really necessary?}}

>>>>

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList} blocks}
\item \tt{\sl{int} size= 1}\textsf{: \#\tt{block}s in \tt{blocks}.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent. Since it is incremented after propagating, it may be behind by 1.}
\item \tt{\sl{int[]} super}\textsf{: \tt{super[i]} stores an approximate index of the superblock of the blocks in \tt{blocks} whose \tt{group} field have value \tt{i}.}
\end{itemize}

@@

@@

\Statex
\Function{Object}{Dequeue()}{}
\State \tt{block newBlock= \Call{new}{\sl{LeafBlock}}} \cmt{Creates a block with null value element, appends it to the tree, computes its order among operations, then computes and returns its response.}
\State \tt{newBlock.element= null}
\State \tt{newBlock.sum\sub{enq}= leaf.blocks[leaf.size].sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq}= leaf.blocks[leaf.size].sum\sub{deq}+1}
\State \tt{leaf.}\Call{Append}{newBlock}
\State \Return\tt{leaf.}\Call{HelpDequeue()}{}
\EndFunction{Dequeue}

@@

\Function{int, int}{FindResponse}{int i, int b}\cmt{Computes the rank  and index of the block in the root of the enqueue that is the response of the \tt{i}th dequeue in the root's \tt{b}th block. Returns \tt{<-1,-->} if the queue is empty.}
\If{\tt{root.blocks[b-1].length + root.blocks[b].num\sub{enq} - i $<$ 0}}
\State \Return \tt{<-1,-->}
\Else
\Statex \cmt{We call the dequeues that return a value \it{non-null dequeues}. $r$th non-null dequeue returns the element of th $r$th enqueue. We can compute \# non-null dequeues in the prefix for a block this way: \#non-null dequeues= length - \#enqueues. Note that the $i$th dequeue in the given block is not a non-null dequeue.}
\State \tt{r\sub{enq}= root.blocks[b-1].sum\sub{enq}- root.blocks[b-1].length + i}
\State \Return \tt{<root.blocks.get(enq, r\sub{enq}).order, r\sub{enq}>}
\EndIf
\EndFunction{FindResponse}

@@

\Function{void}{Propagate()}{}
\If{\bf{not} \Call{Refresh()}{}}
\State \Call{Refresh()}{} \cmt{Lemma Double Refresh}
\EndIf
\If{\tt{this} \bf{is not} \tt{root}}
\State \tt{parent.}\Call{Propagate()}{}
\EndIf
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh()}{}
\State \tt{s= size}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} feild.}
\If{\tt{new.num==0}} \Return{\tt{true}} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, s)}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h+1}} \cmt{Write would work too.}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}}
\EndFor
\State \tt{\Call{CAS}{size, h, h+1}}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{size, h, h+1}} \cmt{Even if another process wins, help to increase the size. The winner might have fallen sleep before increasing \tt{size}.}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Statex \cmt{\textmd{Does binary search for~the value \tt{i} of the given prefix sum \tt{field}. Returns the index of the leftmost block in \tt{blocks[start..end]} whose \sl{field} \tt{f} is $\geq$ \tt{i}}.}
%\State \Return \tt{result block's index}
\EndFunction{BSearch}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\Statex\cmt{Creates a block to be inserted into as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's num\sub{propagated} field. These values are used for incrementing the children's num\sub{propagated} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}
\State \tt{newBlock.order= i}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.size} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{this\sub{dir}= dir.num\sub{propagated}}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

@@

\State \Return \tt{<b, np\sub{left}, np\sub{right}>}
\EndFunction{CreateBlock}

@@

\Statex $\leadsto$ \textsf{Precondition:~\tt{blocks[b].num\sub{enq}$\geq$i}}
\Function{element}{GetEnq}{\sl{int} b, \sl{int} i} 
\If{\tt{this} \bf{is} \tt{leaf}}
\State\Return \tt{blocks[b].element}
\ElsIf{\tt{i $\leq$ blocks[b].num\sub{enq-left}}} \cmt{\tt{i} exists in the left child of this node}
\State \tt{subBlock= left.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{left}+1, blocks[b].end\sub{left}}} \cmt{Search range of left child's subblocks of \tt{blocks[b]}.}
\State \Return\tt{left.}\Call{Get}{i-left.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\Else
\State \tt{i= i-blocks[b].num\sub{enq-left}}
\State\tt{subBlock= right.\Call{BSearch}{sum\sub{enq}, i, blocks[b-1].end\sub{right}+1, blocks[b].end\sub{right}}} \cmt{Search range of right child's subblocks of \tt{blocks[b]}.}
\State \Return\tt{right.}\Call{Get}{i-right.blocks[subBlock-1].sum\sub{enq}, subBlock} 
\EndIf
\EndFunction{GetEnq}

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns the rank of $i$th dequeue in the $b$th block of the node, among the dequeues in the root.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i, super[blocks[b].group]-p, super[blocks[b].group]+p}} 
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}
\If{\tt{dir {\keywordfont is} right}}
%\State \tt{i+= parent.blocks[superBlock-1].sum\sub{deq-right}}
\State \tt{i+= blocks[superBlock].sum\sub{deq-left}} \cmt{consider the dequeues from the right child}
\EndIf
\State \Return\tt{this.parent.}\Call{IndexDeq}{superBlock, i}
\EndIf
\EndFunction{Index}

@@

\Function{void}{Append}{\sl{block} blk} \cmt{Append is only called by the owner of the leaf.}
\State \tt{size+=1} \label{appendEnd} 
\State \tt{blk.group= size} \label{appendStart} 
\State \tt{blocks[size]= blk} 
\State \tt{parent.}\Call{Propagate()}{} 
\EndFunction{Append}

@@

\Function{Object}{HelpDequeue()}{}
\State \tt{<b\sub{deq}, r\sub{deq}>=} \Call{IndexDeq}{leaf.size, 1}\cmt{\tt{r} is the rank among the dequeues of the dequeue of  the \tt{b\sub{deq}}th block in the root containing.}
\State \tt{<r\sub{enq}, r\sub{enq}>=} \Call{FindResponse}{r\sub{deq}, b\sub{deq}} 
 \cmt{\tt{r\sub{enq}} is the rank of the enqueue whose element is the response to the dequeue in the block containing it and \tt{b\sub{deq}} is the index of that block of it in the blocklist. If the response is \tt{null} then r\sub{deq} is -1.} \label{deqRest}
\If{\tt{r\sub{enq}==-1}}
\State \tt{output= null}
\State \tt{root.blocks[b\sub{deq}].num\sub{finished}.inc()} \cmt{shared counter}
\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}}
\State \tt{last\sub{done}= b\sub{deq}}
\EndIf
\pagebreak
\Else
\State \tt{output= \Call{GetEnq}{b\sub{enq}, r\sub{enq}}} \cmt{getting the reponse's \tt{element}.}
\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
\State \tt{root.blocks[b\sub{enq}].num\sub{finished}.inc()}
\If{\tt{root.blocks[b\sub{deq}].num\sub{finished}==root.blocks[b\sub{deq}].num}} 
\State \tt{last\sub{done}= b\sub{deq}}
\ElsIf{\tt{root.blocks[b\sub{enq}].num\sub{finished}==root.blocks[b\sub{enq}].num}}
\State \tt{last\sub{done}= b\sub{enq}}
\EndIf
\EndIf
\State \Return{\tt{output}}
\EndFunction{Dequeue}

@@

\begin{algorithm}
\caption{BlockList}
\begin{algorithmic}[1]
\setcounter{ALG@line}{600}

\Statex \cmt{\textsf{: Supports two operations \tt{blocks.tryAppend(Block b), blocks[i]}. Initially  empty, when \tt{blocks.tryAppend(b, n)} returns true \tt{b} is appended to \tt{blocks[n]} and \tt{blocks[i]} returns $i$th block in the \tt{blocks}. If some instance of \tt{blocks.tryAppend(b, n)} returns \tt{false} there is a concurrent instance of \tt{blocks.tryAppend(b$^\prime$, n)} which has returned \tt{true}.\tt{blocks[0]} contains an empty block with all fields equal to 0 and \tt{end\sub{left}, end\sub{right}} pointers to the first block of the corresponding children.}}

\Statex
\Statex $\diamondsuit$ \tt{\sl{PBRT implementation}}
  \Statex \sf{A persistant red-black tree supporting \tt{append(b, key),get(key=i),split(j)}}.
  \tt{append(b, key) returns \tt{true} in case successful. Since \tt{order, sum\sub{enq}}are both strictly increasing we can use one of them for another.}
\Statex \tt{root: }\sf{pointer to the root of the PBRT}
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} \cmt{\textsf{adds block b to the \tt{root.blocks[n]}}}
\If{\tt{root.size\%$p^2$==0}} \cmt{Help every often $p^2$ operations appended to the root.}
\State \tt{Help()}
\State \tt{CollectGarbage()}
\EndIf
\State \tt{blk.num\sub{finished}= 0}
\State \tt{*oldRoot= \&root.blocks.root}
\State \tt{*newRoot= root.blocks.Append(blk).root}
\State \Return \tt{CAS(root, oldRoot, newRoot)}
\EndFunction{TryAppend}

\Statex
\Statex $\diamondsuit$ \tt{\sl{Array implementation}}
\Statex \tt{blocks[]: }\sf{array of blocks}
\Function{boolean}{TryAppend}{\sl{block} blk, \sl{int} n} 
\State \Return{\tt{CAS(blocks[n], null, blk)}}
\EndFunction{TryAppend}

@@

\Function{void}{Help}{}\cmt{Helps pending operations}
\For{\tt{leaf l} \bf{in leaves}}
\State{\tt{last= l.size-1}}
\cmt{\tt{l.blocks[last]} can not be \tt{null} because size increases after appending, see lines \ref{appendStart}-\ref{appendEnd}.}
\If{\tt{l.blocks[last].element==null}} \cmt{operation is dequeue}
\State \tt{l.blocks[last].response= l.}\Call{HelpDequeue()}{}
\EndIf
\EndFor
\EndFunction{Help}
\Statex

@@

\Function{Block}{FindMostRecentDone}{b}
\For{\tt{leaf l} \bf{in leaves}}
\State\tt{max= Max(l.maxOld, max)}
\EndFor
\State\Return\tt{max} \cmt{This snapshot suffies.}
\EndFunction{FindMostRecentDone}

@@

\Function{response}{FallBack}{op i} \cmt{\it{how to use as exception handling? by adding try catch in all the methods reading the root?}}
